LabMate: a prospectus for types for MATLAB

### 2024-04-11 writing/IMEKOWC2024/LabMate/paper.tex
[ writing ] multiplication - it's the name of the game

Under the hood, LabMate translates selected well behaved MATLAB expressions
to typed expressions in its own core theory of types and values.
%
As matrices feature heavily in MATLAB code, the types of matrices
play a central role in this theory.
%
In particular, we acknowledge that the rows and columns of a matrix might pertain to distinct individual entities, rather as the rows ans columns of a spreadsheet do.
%
Internally, a matrix type looks like
\[\Matrix{R}{C}{E}{\rs}{\cs}
\]
%
We thus \param{} matrix types by two arbitrary `header' types $R$, for the individuals the rows concern, and $C$ for the columns. We may then give two lists, $\rs = [r_1,\ldots,r_m]\in\List{R}$ (so that each $r_i\in R$) and $\cs = [c_1,\ldots,c_n]\in\List{C}$ of those individuals. A \param{}d type $E(r,c)$ gives the type of entry for any combination of row and column individual, with $E(r_i,c_j)$ being the type of the entry $e_{i,j}$. In the diagram below, the $e$s inside the rectangle are the matrix entries computed by the MATLAB code. The $r$s and $c$s outside the rectangle are \remph{meta}data coming from the matrix type, known only to LabMate, and not available as data to MATLAB.
\[\begin{array}{r|ccccc|}
\; & c_1      & \cdots & c_j     & \cdots & c_n     \\
\hline
r_1    & e_{1,1}  &        & e_{1,j} &        & e_{1,n} \\
\vdots &          &        &         &        &         \\
r_i    & e_{i,1}  &        & e_{i,j} &        & e_{i,n} \\
\vdots &          &        &         &        &         \\
r_m    & e_{m,1}  &        & e_{m,j} &        & e_{m,n}  \\
\hline
\end{array}\]
To check that $\Matrix{R}{C}{E}{\rs}{\cs}$ is a valid type, LabMate must:
\begin{enumerate}
\item check that $R$ and $C$ are both valid types;
\item check that $E(r,c)$ is a valid type whenever $r\in R$ and $c\in C$;
\item check that $\rs\in\List{R}$ and $\cs\in\List{C}$.
\end{enumerate}

@@

Firstly, a singleton entry matrix must have singleton header lists $[r]$ and $[c]$.
\[\begin{array}[b]{r|@{}c@{}|l}
\; & c \\
\cline{1-2}
r  & e\in E(r,c) & \quad \in \Matrix{R}{C}{E}{[r]}{[c]} \\
\cline{1-2}
\end{array} 
\]

Secondly, to check a horizontal pasting of matrices, we must ensure that the header and entry types match, that the row headers are the same for left and right, and that the column headers can be split into a prefix for the left matrix and a suffix for the right. In other words, the column headers for the whole are the concatenation ($\append$) of the column headers for the parts.
\[\begin{array}{l}
\begin{array}{r|c|c|}
  & \cs_A & \cs_b \\
\hline
&&\\
&A \in & B \in \\
\rs &\Matrix{R}{C}{E}{\rs}{\cs_A} & \Matrix{R}{C}{E}{\rs}{\cs_B} \\
&&\\
\hline
\end{array}\bigskip\\
\;\;\; \in \Matrix{R}{C}{E}{\rs}{(\cs_A\append\cs_B)}
\end{array}\]

Thirdly, vertical pasting is similar, except that it is the row headers which split into prefix and suffix while the column headers are shared.
\[
\begin{array}[c]{r|c|}
  & \cs \\
  \hline
  \;    & \\
  \;    & A \in \\
  \rs_A & \Matrix{R}{C}{E}{\rs_A}{\cs} \\
  \;    & \\
  \hline
  \;    & \\
  \;    & B \in \\
  \rs_B & \Matrix{R}{C}{E}{\rs_B}{\cs} \\
  \;    & \\
  \hline
\end{array}
\in
\begin{array}{l}
\Matrix{R}{C}{E}{\\ \quad (\rs_A\append\rs_B)}{\cs}
\end{array}
\]

@@

To check if we may add matrices $A$ and $B$, LabMate must:
\begin{enumerate}
\item check that $A$ and $B$ both have the \remph{same} matrix type $\Matrix{R}{C}{E}{\rs}{\cs}$;
\item check that \remph{every} entry type $E(r,c)$ admits addition.
\end{enumerate}
In the short term, LabMate can get by with a hard-coded table of which types admit addition, but more flexible and general approaches to operator overloading (e.g., Haskell's `type classes'~\cite{wadler.blott}) are readily adoptable in due course.

%Matrix types formation can thus be summarised as the inference rule given in Figure~\ref{fig:matrix_intro}.
%\[
%  \infer{\istype{\Matrix{R}{C}{E}{rs}{\cs}}}
%  {
%    \istype{R}
%    &\quad
%    \istype{C}
%    &\quad
%    x : R, y : C \vdash \istype{E(x,y)}
%    &\quad
%    rs : \List{R}
%    &\quad
%    cs : \List{C}
%  }
%\]
%
\begin{figure*}[th]
%\begin{subfigure}{\textwidth}
%  \begin{center}
%    \[
%      \infer{\istype{\Matrix{R}{C}{E}{\rs}{\cs}}}
%      {
%        \istype{R}
%        &\quad
%        \istype{C}
%        &\quad
%        x : R, y : C \vdash \istype{E(x,y)}
%        &\quad
%        \rs : \List{R}
%        &\quad
%        \cs : \List{C}
%      }
%    \]
%  \end{center}
%  \caption{Type formation rule for matrix types}
%  \label{fig:matrix_intro}
%\end{subfigure}
%\hfill
\begin{subfigure}{\textwidth}
  \begin{center}
    \[
      \infer{A + B : \Matrix{R}{C}{E}{\rs}{\cs}}
      {
        x : R, y : C \vdash \isadd{E(x, y)}
        &\quad
        A : \Matrix{R}{C}{E}{\rs}{\cs}
        &\quad
        B : \Matrix{R}{C}{E}{\rs}{\cs}
      }
    \]
  \end{center}
  \caption{Addition rule for matrices}
  \label{fig:matrix_add}
\end{subfigure}
\hfill
\begin{subfigure}{\textwidth}
  \begin{center}
    \[
      \infer{A * B : \Matrix{R}{C}{E''}{\rs}{\cs}}
      {
        \ismult{E}{E'}{E''}
        &\quad
        A : \Matrix{R}{C}{E}{\rs}{\ms}
        &\quad
        B : \Matrix{R}{C}{E'}{\ms}{\cs}
      }
    \]
  \end{center}
  \caption{Multiplication rule for matrices}
  \label{fig:matrix_mul}
\end{subfigure}
% FIXME : \caption is broken
\captionwide{Typing rules for matrices}
\label{fig:matrix_rules}
\end{figure*}

>>>>

Under the hood, LabMate translates selected well behaved MATLAB expressions
to typed expressions in its own core theory of types and values.
%
As matrices feature heavily in MATLAB code, the types of matrices
play a central role in this theory.
%
In particular, we acknowledge that the rows and columns of a matrix might pertain to distinct individual entities, rather as the rows ans columns of a spreadsheet do.
%
Internally, a matrix type looks like
\[\Matrix{R}{C}{E}{\rs}{\cs}
\]
%
We thus \param{} matrix types by two arbitrary `header' types $R$, for the individuals the rows concern, and $C$ for the columns. We may then give two lists, $\rs = [r_1,\ldots,r_m]\in\List{R}$ (so that each $r_i\in R$) and $\cs = [c_1,\ldots,c_n]\in\List{C}$ of those individuals. A \param{}d type $E(r,c)$ gives the type of entry for any combination of row and column individual, with $E(r_i,c_j)$ being the type of the entry $e_{i,j}$. In the diagram below, the $e$s inside the rectangle are the matrix entries computed by the MATLAB code. The $r$s and $c$s outside the rectangle are \remph{meta}data coming from the matrix type, known only to LabMate, and not available as data to MATLAB.
\[\begin{array}{@{}r|ccccc|}
\; & c_1      & \cdots & c_j     & \cdots & c_n     \\
\hline
r_1    & e_{1,1}  &        & e_{1,j} &        & e_{1,n} \\
\vdots &          &        &         &        &         \\
r_i    & e_{i,1}  &        & e_{i,j} &        & e_{i,n} \\
\vdots &          &        &         &        &         \\
r_m    & e_{m,1}  &        & e_{m,j} &        & e_{m,n}  \\
\hline
\end{array}\]
To check that $\Matrix{R}{C}{E}{\rs}{\cs}$ is a valid type, LabMate must:
\begin{enumerate}
\item check that $R$ and $C$ are both valid types;
\item check that $E(r,c)$ is a valid type whenever $r\in R$ and $c\in C$;
\item check that $\rs\in\List{R}$ and $\cs\in\List{C}$.
\end{enumerate}

@@

Firstly, a singleton entry matrix must have singleton header lists $[r]$ and $[c]$.
\[\begin{array}[b]{@{}r|@{}c@{}|l}
\; & c \\
\cline{1-2}
r  & e\in E(r,c) & \quad \in \Matrix{R}{C}{E}{[r]}{[c]} \\
\cline{1-2}
\end{array} 
\]

Secondly, to check a horizontal pasting of matrices, we must ensure that the header and entry types match, that the row headers are the same for left and right, and that the column headers can be split into a prefix for the left matrix and a suffix for the right. In other words, the column headers for the whole are the concatenation ($\append$) of the column headers for the parts.
\[\begin{array}{l}
\begin{array}{@{}r|c|c|}
  & \cs_A & \cs_b \\
\hline
&&\\
&A \in & B \in \\
\rs &\Matrix{R}{C}{E}{\rs}{\cs_A} & \Matrix{R}{C}{E}{\rs}{\cs_B} \\
&&\\
\hline
\end{array}\bigskip\\
\;\;\; \in \Matrix{R}{C}{E}{\rs}{(\cs_A\append\cs_B)}
\end{array}\]

Thirdly, vertical pasting is similar, except that it is the row headers which split into prefix and suffix while the column headers are shared.
\[
\begin{array}[c]{@{}r|c|}
  & \cs \\
  \hline
  \;    & \\
  \;    & A \in \\
  \rs_A & \Matrix{R}{C}{E}{\rs_A}{\cs} \\
  \;    & \\
  \hline
  \;    & \\
  \;    & B \in \\
  \rs_B & \Matrix{R}{C}{E}{\rs_B}{\cs} \\
  \;    & \\
  \hline
\end{array}
\in
\begin{array}{l}
\Matrix{R}{C}{E}{\\ \quad (\rs_A\append\rs_B)}{\cs}
\end{array}
\]

@@

To check if we may add matrices $A$ and $B$, LabMate must:
\begin{enumerate}
\item check that $A$ and $B$ both have the \remph{same} matrix type $\Matrix{R}{C}{E}{\rs}{\cs}$;
\item check that \remph{every} entry type $E(r,c)$ admits addition.
\end{enumerate}
In the short term, LabMate can get by with a hard-coded table of which types (e.g. \texttt{double}) admit addition, but more flexible and general approaches to operator overloading (e.g., Haskell's `type classes'~\cite{wadler.blott}) are readily adoptable in due course.

Multplication is rather more involved, of course. The usual requirement that the matrix sizes `meet in the middle' generalises straightforwardly to lists of headers,
\[\begin{array}{@{}c@{}}
\begin{array}[c]{@{}r|c|}
  \;  & \ms \\
  \hline
  \;  &     \\
  \;  &  A \in \\
  \rs & \Matrix{R}{M}{E_1}{\rs}{\ms} \\
  \;  & \\
  \hline
\end{array}
  \times
\begin{array}[c]{@{}r|c|}
  \;  & \cs \\
  \hline
  \;  &     \\
  \;  &     \\
  \;  &  B \in \\
  \ms & \Matrix{M}{C}{E_2}{\\ & \ms}{\cs} \\
  \;  &     \\
  \;  & \\
  \hline
\end{array}
\bigskip\\
\in \qquad
\begin{array}[c]{@{}r|c|}
  \;  & \cs \\
  \hline
  \;  &     \\
  \rs & \Matrix{R}{C}{E_3}{\\ & \rs}{\cs} \\
  \;  & \\
  \hline
\end{array}
\end{array}
\]
but we must also check that the entry types are compatible. Specifically, LabMate must:
\begin{enumerate}
  \item check, for all $r$, $m$ and $c$, that if $a\in E_1(r,m)$ and $b\in E_2(m,c)$, then
    $a\times b \in E_3(r,c)$;
  \item check that \remph{every} entry type $E_3(r,c)$ admits addition.
\end{enumerate}

These conditions are straightforward in the special case where $R = C = \One$ and the $E$s are standard numerical types. There are, however, more motivating possibilities.

%Matrix types formation can thus be summarised as the inference rule given in Figure~\ref{fig:matrix_intro}.
%\[
%  \infer{\istype{\Matrix{R}{C}{E}{rs}{\cs}}}
%  {
%    \istype{R}
%    &\quad
%    \istype{C}
%    &\quad
%    x : R, y : C \vdash \istype{E(x,y)}
%    &\quad
%    rs : \List{R}
%    &\quad
%    cs : \List{C}
%  }
%\]
%
\begin{figure*}[th]
%\begin{subfigure}{\textwidth}
%  \begin{center}
%    \[
%      \infer{\istype{\Matrix{R}{C}{E}{\rs}{\cs}}}
%      {
%        \istype{R}
%        &\quad
%        \istype{C}
%        &\quad
%        x : R, y : C \vdash \istype{E(x,y)}
%        &\quad
%        \rs : \List{R}
%        &\quad
%        \cs : \List{C}
%      }
%    \]
%  \end{center}
%  \caption{Type formation rule for matrix types}
%  \label{fig:matrix_intro}
%\end{subfigure}
%\hfill
%\begin{subfigure}{\textwidth}
%  \begin{center}
%    \[
%      \infer{A + B : \Matrix{R}{C}{E}{\rs}{\cs}}
%      {
%        x : R, y : C \vdash \isadd{E(x, y)}
%        &\quad
%        A : \Matrix{R}{C}{E}{\rs}{\cs}
%        &\quad
%        B : \Matrix{R}{C}{E}{\rs}{\cs}
%      }
%    \]
%  \end{center}
%  \caption{Addition rule for matrices}
%  \label{fig:matrix_add}
%\end{subfigure}
%\hfill
\begin{subfigure}{\textwidth}
  \begin{center}
    \[
      \infer{A * B : \Matrix{R}{C}{E''}{\rs}{\cs}}
      {
        \ismult{E}{E'}{E''}
        &\quad
        A : \Matrix{R}{C}{E}{\rs}{\ms}
        &\quad
        B : \Matrix{R}{C}{E'}{\ms}{\cs}
      }
    \]
  \end{center}
  \caption{Multiplication rule for matrices}
  \label{fig:matrix_mul}
\end{subfigure}
% FIXME : \caption is broken
\captionwide{Typing rules for matrices}
\label{fig:matrix_rules}
\end{figure*}
